Correlation tests

### 2025-04-03 analysis/manuscript.tex
Proofreading

hoaddata, developed and maintained by the author to support open access
monitoring and research (Jahn, 2025), is an R data package that
regularly collects information on hybrid open access uptake from
multiple openly available data sources (Jahn, 2024). It combines
article-level metadata from Crossref (Hendricks et al., 2020) and
affilation metadata from OpenAlex (Priem et al., 2022) with
transformative agreement information from Transformative Agreement Data
dump used by the cOAlition S Journal Checker Tool (JCT)\footnote{\url{https://journalcheckertool.org/transformative-agreements/}},
which links journal and institutional data about participating research
organisations to agreements in the ESAC registry.

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). Web of Science is a selective, base
research-focused database with dedicated panels to choose journals for
inclusion (Stahlschmidt \& Stephen, 2022; Visser et al., 2021). The
collections considered in this study were the Science Citation Index
Expanded (SCIE), the Social Sciences Citation Index (SSCI) and the Arts
\& Humanities Citation Index (AHCI).

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}},
as provided by hoaddata. The JCT journal data in hoaddata were enriched
with ISSN variants linked to an ISSN-L. To identify hybrid journals, a
comprehensive exclusion of fully open access journals was performed
using multiple journal lists (DOAJ, OpenAlex, Bielefeld). These data
were used to determine the publication volume of hybrid journals for
each database independently.

hoaddata relies on Crossref for obtaining journal publication volume and
open access status through Creative Commons licence information relative
to the published version (``version of record''). The article metadata
included DOIs, publication dates, open access information as well as
author roles and institutional affiliations. Publication years were
determined using the earliest known date of publication in a journal. In
hoaddata, this corresponded to Crossref's issued date. For Web of
Science and Scopus, the earliest publication date was used where
available, with Scopus dates specifically determined by the KB through
version tracking of the raw data.

@@

Articles in hybrid journals were considered open access when they were
made freely available under a Creative Commons license on publishers'
platforms. While hoaddata sourced this information from Crossref license
metadata, Web of Science and Scopus relied on Unpaywall as evidence
source. Unpaywall also uses Crossref license metadata, but supplements
them by parsing publisher websites directly, addressing cases where
publishers do not provide machine-readable Creative Commons license
information (Piwowar et al., 2018). This additional parsing remains
necessary despite transformative agreement workflows recommend the
depositing license information during DOI registration (Geschuhn \&
Stone, 2017). Both Web of Science and Scopus defined hybrid open access
consistently as content available under Creative Commons licenses on
publisher platforms according to their
documentations\footnote{\url{https://webofscience.help.clarivate.com/en-us/Content/open-access.html}}\footnote{\url{https://blog.scopus.com/posts/scopus-filters-for-open-access-type-and-green-oa-full-text-access-option}},
distinguishing it from bronze open access that lack such explicit
license information, or use publisher-specific licenses.

@@

Because neither Web of Science nor Scopus support ROR IDs, the
institution identifier used by the JCT, a two-step matching process was
implemented to harmonise affiliation data. First, 2,782,540 articles
from 6,457 institutions with ROR-IDs in the JCT data since 2017
(according to hoaddata) were processed to map first authors' ROR-IDs to
corresponding proprietary affiliation identifier in Web of Science and
Scopus using DOI matching. Then, an algorithm selected the most frequent
ROR ID and proprietary identifier pairs to handle multiple affiliations
and organisational hierarchy differences.

This process linked 6,375 ROR IDs to 4,894 Scopus Affiliation IDs, and
6,034 ROR IDs to 2,422 enhanced affiliation strings in the Web of
Science. Quality evaluation through random sampling of 50 pairs revealed
an error rate of 22\% for Web of Science (11 mismatches) and 6\% for
Scopus (3 mismatches). Upon inspection, these mismatches primarily
occurred with less-represented institutions having only a few
publications, introduced through multiple affiliations of single
authors. The difference between databases suggests that Scopus's
affiliation control aligns more closely with ROR than that of the Web of
Science.

@@

Based on these generated matching tables, articles eligible under
transformative agreements could also be retrieved from Web of Science
and Scopus, although they did not contain the ROR IDs used by the JCT.
The estimation of eligible articles followed Jahn (2025) and included a
matching of both journals and participating institutions. The matching
also took into account the duration of agreements according to the ESAC
registry, with only those matches where an agreement was actually in
place being considered for subsequent analysis. A related study (Jonge
et al., 2025), applied to publications funded by the Dutch Research
Council (NWO) and validated against internal invoicing data, confirmed
that such matching can accurately identify most articles under
transformative agreements.

@@

\caption{\label{fig-upset_coverage_results}Upset graph}

@@

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C illustrates the distribution for each
combination. It shows that there is a large spread across the journals
covered by all three data sources. Furthermore, journals in this set
published more on average than journals covered by only one or two data
sources. In particular, the journals covered exclusively by hoaddata
were substantially smaller than those covered by all three sources. Upon
inspection, these were often newly launched hybrid journals, which
explains the relatively low five-year publication volume. An example is
\emph{Digital society} that published 86 articles. This hybrid journal
was launched in 2022, being covered by various Springer Nature
transformative agreements since then.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 with a focus on publisher portfolios. The analysis
highlights the dominance of the three largest publishers--Elsevier,
Springer Nature, and Wiley--, which collectively accounted for 47\% of
hybrid journals and 62\% of articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358 articles
(33\% of the total) published across 1,951 hybrid journals (16\% of the
total). Springer Nature followed with 1,247,578 articles (17\%) in
2,311, although recording the largest number of hybrid journals (19\%).
Wiley accounted for 858,939 articles (12\%) in 1,382 hybrid journals
(11\%). The remaining 54 publishers collectively accounted for 2,749,847
articles (38\%) in 6,476 hybrid journals (53\%).

@@

\caption{\label{fig-upset_coverage_results_publisher}Upset graph
publisher}

@@

The three largest publishers--Elsevier, Springer Nature and Wiley--were
best represented in the exclusive intersection of all three data sources
(hoaddata, Scopus, and Web of Science). Together, they comprised 4,384
hybrid journals (55\% of the intersectional set) and dominated article
coverage (n = 4,174,315; 66\%), as determined through shared DOIs. When
examining publication volume per journal, (Figure
\ref{fig-upset_coverage_results_publisher}C), Elsevier published, on
average, the largest journals, followed by Springer Nature and Wiley.

@@

Examining publisher portfolios not covered by hoaddata but present in
Scopus or Web of Science identified several publishers with missing CC
license information in Crossref. In particular, Emerald represented 322
journals with 86,409 articles, AIP Publishing accounted for 24 journals
with 64,898 articles, and World Scientific recorded 87 journals and
42,531 articles. In total, 9 publishers were not represented in
Crossref.

@@

\caption{\label{fig-uptake_overview}Open Access indicato}

@@

Figure \ref{fig-uptake_overview}A shows a moderate growth of open access
in hybrid journals, which is consistent across hoaddata (black line),
Scopus (orange line), and Web of Science (blue line). According to
hoaddata, hybrid open access uptake increased from \% (n = ) in 2019 to
\% (n = ) in 2023. Similarly, Scopus recorded an growth from 6.5\% (n =
84,648) in 2019 to 19\% (n = 322,850) in 2023. However, Web of Science
recorded higher open access uptake in early years, before converging in
2023, from 11\% (n = 137,202) in 2019 to 20\% (n = 255,481) in 2023,
suggesting an different approach towards labeling hybrid open access by
the Web of Science.

Similarly, hybrid open access by transformative agreements increased
between 2019 and 2023 (Figures \ref{fig-uptake_overview}B and C). Trends
were consistent for first (dashed line) and corresponding author (dotted
line) affiliations. According to Scopus, 479,297 open access articles
could be attributed to transformative agreements based on first author
metadata (increasing from 15,341 to 206,084) and 489,262 using
corresponding author metadata (from 15,444 to 210,816). Web of Science
recorded 493,028 open access articles via transformative agreements
using first author metadata (increasing from 21,871 to 174,126) and
500,076 using corresponding author metadata (from 22,092 to 177,030).
hoaddata, lacking corresponding author data, linked 0 articles to
transformative agreements but showed slower growth than Scopus and Web
of Science, from 0 in 2019 to 0 in 2023.

From 2021 (hoadata, Scopus) resp. 2022 (Web of Science), transformative
agreements enabled the majority of hybrid open access. For first
authors, the share ranged between \% (hoaddata), 64\% (Scopus), and 68
\% (Web of Science) in 2023. For corresponding authors, the shares were
slightly larger, with Scopus recording 65\% and Web of Science 69\% in
2023. However, substantial hybrid open access was still facilitated
outside transformative agreements, likely through APCs paid from
discretionary research funds (Jahn et al., 2022; Suber, 2012).

@@

\caption{\label{fig-uptake_publisher}Open Access indicato}

@@

When comparing countries, consistent patterns were observed across data
sources for the five-year period 2019 to 2023. Figure
\ref{fig-uptake_country} presents hybrid open access indicators by
country, comparing hoaddata (x-axis) with Web of Science and Scopus
(y-axis). Indicators calculated from these proprietary databases are
shown for both first and corresponding authors, with full counting used
to account for multiple country affiliations.

@@

\caption{\label{fig-uptake_country}Open Access by country}

@@

A closer examination of the proportion of hybrid open access enabled by
transformative agreements (see Figure \ref{fig-uptake_country}D)
revealed that the French Elsevier agreement on delayed open access was
not included in JCT data, as it did not qualify as a transformative
agreement. Medium-sized European countries again showed a high
proportion of hybrid open access through transformative agreements,
highlighting the impact of this licensing model across all three data
sources. In contrast, the United States had a low proportion of hybrid
open access enabled by agreements, suggesting that a substantial number
of open access articles were likely financed through other means, such
as article processing charges (APCs).

@@

This study of over 13,000 hybrid journals included in the cOAlition
Journal Checker Tool shows a rise in open access due to transformative
agreements between 2019 and 2023, although most articles remained
paywalled. While transformative agreements accounted for the majority of
open access, many articles continue to become open through the payment
of individual publication fees. Hybrid open access and transformative
agreements remain concentrated among a small group of large commercial
publishers, with high-income European countries---alongside South
Africa---showing high adoption rates. In contrast, the three most
productive countries, China, the United States, and India, were ranked
much behind other countries in the adoption of transformative
agreements. These trends were consistent across the investigated open
data source hoaddata, derived from Crossref and OpenAlex, and the
established proprietary bibliometric databases Scopus and Web of
Science. An open question remains whether this uneven distribution
reflects temporary implementation gaps, inherent inequities in the
transformative agreement model, or countries voluntarily refraining from
such agreements.

Comparing the data sources examined, the study results show strong
correlations by country affiliation, despite differences in journal and
article coverage, and in the availability of metadata on corresponding
authorship and open access status. However, several notable
discrepancies warrant discussion. They arise from limitations in both
open and proprietary databases.

The coverage analysis reveals that hybrid journals are well indexed in
all three data sources, particularly in terms of article coverage.
Differences can be found for journals targeting practitioners or local
non-English language communities, with many such titles indexed
exclusively in Crossref and Scopus. Using the open index Crossref as
bibliometric database in hoaddata demonstrated particular strength in
identifying newly established hybrid journals, a notable finding given
that transformative agreements primarily target existing
subscription-based journals.

@@

A notable advantage of using Scopus and Web of Science to study the
impact of transformative agreements is their comprehensive
differentiation of journal articles by document type, as many agreements
restrict funding to original articles and reviews. This resulted in
slightly higher open access rates in Scopus and Web of Science.

A similar limitation of studying open access with less selective
databases is the often reported lack of corresponding authorship
information (Fraser et al., 2023; Haucap et al., 2021; Shu \& Larivière,
2023). However, indicators based on first authors, which have often been
used as a proxy for determining open access funding, and corresponding
authors show a high level of correlation, in line with previous
research, reflecting disciplinary norms in scholarly publishing with
regard to contributorship and author roles and positions (Larivière et
al., 2016). First authors are often considered to have carried out the
main research underlying a paper, while the corresponding author is
often recognised as having supervised the research (Fox et al., 2018;
Mattsson et al., 2010). It is therefore not surprising that both are
strongly correlated, suggesting that first and corresponding authors
share the same country affiliation. Moreover, in most cases the first
author is indeed the same as the corresponding author
(Chinchilla-Rodríguez et al., 2024).

The slight increase in the number of open access articles by
corresponding authors compared to the number of first authors in the
last two years observed across Springer Nature hybrid journals raises
the question about institutional open access sponsorship practices and
if they influence the assignment of author roles among co-authors,
especially when funding opportunities are not consistently available
within the author teams (Gumpenberger et al., 2018). Previous research
notes effects of institutionalised bibliometrics praticses on the
valuations of authorship positions (Helgesson, 2020), although, to my
knowledge, no such evidence was reported for open access funding yet. As
such, first authorship can reliably be used in large-scale studies and
those focusing on national consortia to assess the impact of
transformative agreements.

While OpenAlex's affiliation coverage is less complete, it shows strong
correlations with Scopus (Akbaritabar et al., 2024; Alperin et al.,
2024). In this study, this limitation likely reduced the number of
articles confidently attributed to transformative agreements, despite
high correlations between OpenAlex, Scopus, and Web of Science at the
country level. Nevertheless, OpenAlex's native ROR ID integration offers
a distinct advantage: unlike Scopus and Web of Science, which require
reconciliation of proprietary organisation identifiers with this open
standard, it enables more reliable identification of agreement-enabled
articles.

Comparing scholarly databases helps to identify discrepancies in open
access evidence. Not all publishers deposit CC licences with Crossref,
but Unpaywall, integrated into Scopus and Web of Science, can make this
transparent, because the open access discovery services also analyse
publishers' websites. However, inconsistencies remain: Web of Science
incorrectly identifies Elsevier's delayed open access as hybrid, while
Scopus correctly reflects it. Unpaywall also missed a substantial number
of open access articles with CC license according to Crossref, resulting
in fewer open access articles in Scopus and Web of Science, probably due
to problems parsing the publisher's website. As these differences may
affect comparability between databases, research and monitoring
exercises should not rely on a single source of evidence. Instead, it is
recommended that the choice of open access evidence sources should be
documented and verified by combining multiple data sources (Huang et
al., 2020). It is particularly useful to include expert-curated lists of
journals to avoid misclassifications based on a journal's business model
(Visser et al., 2021).

@@

@@

>>>>

hoaddata, developed and maintained by the author to support open access
monitoring and research (e.g. Jahn (2025)), is an R data package that
regularly collects information on hybrid open access uptake from
multiple openly available data sources (Jahn, 2024). It combines
article-level metadata from Crossref (Hendricks et al., 2020) and
affilation metadata from OpenAlex (Priem et al., 2022) with
transformative agreement information from the Transformative Agreement
Data dump used by the cOAlition S Journal Checker Tool (JCT)\footnote{\url{https://journalcheckertool.org/transformative-agreements/}},
which links journal and institutional data about participating research
organisations to agreements in the ESAC registry.

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). Web of Science is a selective, base
research-focused database (Stahlschmidt \& Stephen, 2022; Visser et al.,
2021). The collections considered in this study were the Science
Citation Index Expanded (SCIE), the Social Sciences Citation Index
(SSCI) and the Arts \& Humanities Citation Index (AHCI).

@@

Following Jahn (2025), the starting point was a unified dataset of
several safeguarded JCT snapshots\footnote{\url{https://github.com/njahn82/jct_data}}.
The JCT journal data were enriched with ISSN variants linked to an
ISSN-L. To identify hybrid journals, a comprehensive exclusion of fully
open access journals was performed using multiple journal lists (DOAJ,
OpenAlex, Bielefeld). These data were made available via hoaddata and
used to determine the publication volume of hybrid journals for each
database independently.

hoaddata relies on Crossref for obtaining journal publication volume and
open access status through Creative Commons (CC) licence information
relative to the published version (``version of record''). The article
metadata included DOIs, publication dates, open access information as
well as author roles and affiliations. Publication years were determined
using the earliest known date of publication in a journal. In hoaddata,
this corresponded to Crossref's issued date. For Web of Science and
Scopus, the earliest publication date was used where available, with
Scopus dates specifically determined by the KB through version tracking
of the raw data.

@@

Articles in hybrid journals were considered open access when they were
made freely available under a CC license on publisher platforms. While
hoaddata obtained this information from Crossref license metadata, Web
of Science and Scopus relied on Unpaywall as evidence source. Unpaywall
also uses Crossref license metadata, but supplements them by parsing
publisher websites directly, addressing cases where publishers do not
provide machine-readable CC license information (Piwowar et al., 2018).
This additional parsing remains necessary despite transformative
agreement workflows recommend to deposit CC licenses during DOI
registration (Geschuhn \& Stone, 2017). Both Web of Science and Scopus
defined hybrid open access consistently as content available under CC
licenses on publisher platforms according to their
documentations\footnote{\url{https://webofscience.help.clarivate.com/en-us/Content/open-access.html}}\footnote{\url{https://blog.scopus.com/posts/scopus-filters-for-open-access-type-and-green-oa-full-text-access-option}},
distinguishing it from bronze open access that lack such explicit
license information, or use publisher-specific licenses (Piwowar et al.,
2018).

@@

Because neither Web of Science nor Scopus support ROR at the time of the
data retrieval, the institution identifier used by the JCT, a two-step
matching process was implemented to harmonise affiliation data. First,
2,782,540 articles from 6,457 institutions with ROR-IDs in the JCT data
since 2017 (according to hoaddata) were processed to map first authors'
ROR-IDs to corresponding proprietary affiliation identifier in Web of
Science and Scopus using DOI matching. Then, an algorithm selected the
most frequent ROR ID and proprietary identifier pairs to handle multiple
affiliations and organisational hierarchy differences.

This process linked 6,375 ROR IDs to 4,894 Scopus Affiliation IDs, and
6,034 ROR IDs to 2,422 enhanced affiliation strings in the Web of
Science. Quality evaluation through random sampling of 50 pairs revealed
an error rate of 22\% for Web of Science (11 mismatches) and 6\% for
Scopus (3 mismatches). Upon inspection, these mismatches primarily
occurred with less-represented institutions having only a few
publications, introduced through multiple affiliations of single
authors. The difference between databases suggests that Scopus'
affiliation control aligns more closely with ROR than that of the Web of
Science.

@@

Based on these so-compiled matching tables, articles eligible under
transformative agreements could also be obtained from Web of Science and
Scopus, although they did not contain the ROR IDs used by the JCT. The
estimation of eligible articles followed Jahn (2025) and included a
matching of both journals and participating institutions according to
the JCT. The matching also took into account the duration of agreements
according to the ESAC registry, with only those matches where an
agreement was actually in place being considered for subsequent
analysis. A related study (Jonge et al., 2025), applied to publications
funded by the Dutch Research Council (NWO) and validated against
internal invoicing data, confirmed that such matching can accurately
identify most articles under transformative agreements.

@@

\caption{\label{fig-upset_coverage_results}Comparison of the hybrid open
access indexing coverage of hoaddata (based on Crossref), Scopus and Web
of Science, 2019-2023. Only hybrid journals present in the cOAlition S
Transformative Agreement Data dump with at least one open access article
are considered. A) presents the number of journals, B) the number of
articles (DOI), distinguishing between shared DOI corpus and surplus in
hoaddata. Box plots (C) shows the five-years journal article volume
(log-scale).}

@@

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C illustrates the distribution for each
combination. It shows a large spread across the journals jointly covered
by all three data sources. These journals published more on average than
journals covered by only one or two data sources. In particular, the
journals covered exclusively by hoaddata were substantially smaller.
Upon inspection, these were often newly launched hybrid journals, which
explains the relatively low five-year publication volume. An example is
\emph{Digital society} that published 86 articles. This hybrid journal
was launched in 2022, being covered by various Springer Nature
transformative agreements since then.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 with a focus on publisher portfolios. The analysis
highlights the dominance of the three largest publishers, Elsevier,
Springer Nature, and Wiley, which collectively accounted for 47\% of
hybrid journals and 62\% of articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358 articles
(33\% of the total) published across 1,951 hybrid journals (16\% of the
total). Springer Nature followed with 1,247,578 articles (17\%) in
2,311, although recording the largest number of hybrid journals (19\%).
Wiley accounted for 858,939 articles (12\%) in 1,382 hybrid journals
(11\%). The remaining 54 publishers collectively accounted for 2,749,847
articles (38\%) in 6,476 hybrid journals (53\%).

@@

\caption{\label{fig-upset_coverage_results_publisher}Comparison of the
hybrid open access indexing coverage of hoaddata (based on Crossref),
Scopus and Web of Science, 2019-2023 by publisher. Only hybrid journals
present in the cOAlition S Transformative Agreement Data dump with at
least one open access article are considered. A) presents the percentage
of journals by publisher, B) the number of articles by publisher (shared
DOI). Box plots (C) shows the five-years journal article volume
(log-scale) by publisher.}

@@

The three largest publishers, Elsevier, Springer Nature, and Wiley, were
best represented in the exclusive intersection of all three data sources
(hoaddata, Scopus, and Web of Science). Together, they comprised 4,384
hybrid journals (55\% of the intersectional set) and dominated article
coverage (n = 4,174,315; 66\%), as determined through shared DOIs. When
examining publication volume per journal, (Figure
\ref{fig-upset_coverage_results_publisher}C), Elsevier published, on
average, the largest journals, followed by Springer Nature and Wiley.

@@

Examining publisher portfolios not covered by hoaddata but present in
Scopus or Web of Science identified several publishers with missing CC
license information in Crossref. In particular, Emerald represented 322
journals with 86,409 articles, AIP Publishing accounted for 24 journals
with 64,898 articles, and World Scientific recorded 87 journals and
42,531 articles. In total, 9 publishers did not share CC licenses with
Crossref and were therefore not represented in hoaddata.

@@

\caption{\label{fig-uptake_overview}Development of open access in hybrid
journals included in transformative agreements, 2019 and 2023 by data
source and author role. Figure shows the open access percentage (A), the
number (B) and percentage (C) of open access articles enabled
transformative agreements.}

@@

Figure \ref{fig-uptake_overview}A shows a moderate growth of open access
in hybrid journals, which is consistent across hoaddata (black line),
Scopus (yellow line), and Web of Science (blue line). According to
hoaddata, hybrid open access uptake increased from 6.4\% (n = 85,071) in
2019 to 19\% (n = 302,358) in 2023. Similarly, Scopus recorded an growth
from 6.5\% (n = 84,648) in 2019 to 19\% (n = 322,850) in 2023. However,
Web of Science recorded higher open access uptake in early years, before
converging in 2023, from 11\% (n = 137,202) in 2019 to 20\% (n =
255,481) in 2023, suggesting an different approach towards labeling
hybrid open access by the Web of Science.

Similarly, hybrid open access by transformative agreements increased
between 2019 and 2023 (Figures \ref{fig-uptake_overview}B and C). Trends
were consistent when measuring first (dashed line) and corresponding
author (dotted line) affiliations. According to Scopus, 479,297 open
access articles could be attributed to transformative agreements based
on first author metadata (increasing from 15,341 to 206,084) and 489,262
using corresponding author metadata (from 15,444 to 210,816). Web of
Science recorded 493,028 open access articles via transformative
agreements using first author metadata (increasing from 21,871 to
174,126) and 500,076 using corresponding author metadata (from 22,092 to
177,030). hoaddata, lacking corresponding author data, linked 0 articles
to transformative agreements but showed slower growth than Scopus and
Web of Science, from 0 in 2019 to 0 in 2023.

From 2021 (hoadata, Scopus) resp. 2022 (Web of Science), transformative
agreements enabled the majority of hybrid open access. For first
authors, the share ranged between 61\% (hoaddata), 64\% (Scopus), and 68
\% (Web of Science) in 2023. For corresponding authors, the shares were
slightly larger, with Scopus recording 65\% and Web of Science 69\% in
2023. However, substantial hybrid open access was still facilitated
outside transformative agreements, likely through APCs paid from
discretionary research funds (Jahn et al., 2022; Suber, 2012).

@@

\caption{\label{fig-uptake_publisher}Development of open access in
hybrid journals included in transformative agreements, 2019 and 2023, by
data source, author role and publisher. Figure shows the open access
percentage (A), the number (B) and percentage (C) of open access
articles enabled transformative agreements.}

@@

When comparing countries, consistent patterns were observed across data
sources for the five-year period 2019 to 2023. Figure
\ref{fig-uptake_country} presents hybrid open access indicators by
country, comparing hoaddata (x-axis) with Web of Science and Scopus
(y-axis). Indicators calculated from these proprietary databases are
shown for both first and corresponding authors, with full counting used
to account for multiple country affiliations (Hottenrott et al., 2021).

@@

\caption{\label{fig-uptake_country}Comparison of hybrid open access by
country, 2019-2023. Scatterplots distinguish between proprietary
database Scopus and Web of Science, and author role. x axis shows
hoaddata indicators. A) Five-years article volume, B) open access
percentage in hybrid journals, (A and B limited to countries with
\textgreater{} 10.000 articles published), C) number and D) percentage
of open access articles enabled by transformative agreements (limited to
countries with \textgreater{} 1.000 open access articles enabled by
transformative agreements). Line represents line of equality. An
interactive version is accessible here:
\url{https://najkoja-hoa-replication.hf.space/}}

@@

Medium-sized European countries again showed a high proportion of hybrid
open access through transformative agreements (see Figure
\ref{fig-uptake_country}D), highlighting the impact of this licensing
model across all three data sources. In contrast, the United States had
a low proportion of hybrid open access enabled by agreements, suggesting
that a substantial number of open access articles from US-based authors
were likely financed through other means.

@@

This study of over 13,000 hybrid journals included in the Transformative
Agreement Data underlying the cOAlition S Journal Checker Tool shows a
rise in open access due to transformative agreements between 2019 and
2023, although most articles remained paywalled. While transformative
agreements accounted for the majority of open access, many articles
continue to become open through the payment of individual publication
fees. Hybrid open access and transformative agreements remain
concentrated among a small group of large commercial publishers, with
high-income European countries---alongside South Africa---showing high
adoption rates. In contrast, the three most productive countries, China,
the United States, and India, were ranked much behind other countries in
the adoption of transformative agreements. Open questions remain whether
this uneven distribution reflects temporary implementation gaps,
inherent inequities in the transformative agreement model, or countries
voluntarily refraining from such agreements.

These trends were consistent across the investigated open data source
hoaddata, derived from Crossref and OpenAlex, and the established
proprietary bibliometric databases Scopus and Web of Science. Notably,
the results show strong correlations by country affiliation, aligning
with previous studies (Akbaritabar et al., 2024; Alperin et al., 2024),
which supports the use of open research information for large-scale
analyses of hybrid open access. However, the observed differences in
coverage of journals, articles, and metadata availability warrant
discussion, affecting not only open data sources but also proprietary
databases when used in isolation.

To begin with, the coverage analysis reveals that hybrid journals are
well indexed in all three data sources, particularly in terms of article
coverage. Differences can be found for journals targeting practitioners
or local non-English language communities, with many such titles indexed
exclusively in Crossref and Scopus. Using the open index Crossref as a
bibliometric database in hoaddata demonstrated particular strength in
identifying newly established hybrid journals, a notable finding given
that transformative agreements primarily target existing
subscription-based journals.

@@

One often reported limitation of studying open access with less
selective databases is the lack of corresponding authorship information
(Fraser et al., 2023; Haucap et al., 2021; Shu \& Larivière, 2023).
However, indicators based on first authors, which have often been used
as a proxy for determining open access funding, and corresponding
authors show a high level of correlation, reflecting disciplinary norms
in scholarly publishing with regard to contributorship and author roles
and positions (Larivière et al., 2016). First authors are often
considered to have carried out the main research underlying a paper,
while the corresponding author is often recognised as having supervised
the research (Fox et al., 2018; Mattsson et al., 2010). Unsurprisingly,
measures based on first or corresponding authorships are strongly
correlated, suggesting that these authors share the same country
affiliation. Moreover, in most cases, the first author is identical the
corresponding author (Chinchilla-Rodríguez et al., 2024). Thus, first
authorship can reliably be used in large-scale studies to assess the
impact of transformative agreements.

Despite this correlation, the recent slight increase in open access
articles by corresponding authors over first authors across Springer
Nature hybrid journals prompts questions about how institutional open
access sponsorship practices influence author roles and assignments
within co-author teams, especially as funding opportunities vary
(Gumpenberger et al., 2018). Previous research has highlighted how
institutionalized bibliometric practices can affect the valuation of
authorship positions (Helgesson, 2020), underscoring the need to be
cautious of monetary authorship influences (Maddi \& Silva, 2024).

Affiliation data is another important data element for estimating open
access enabled by transformative agreements. Although OpenAlex's
affiliation coverage is less comprehensive, which likely reduced the
number of articles confidently attributed to transformative agreements
in this study, it still shows high correlations with Scopus and Web of
Science at the country level. However, OpenAlex's native ROR ID
integration offers a distinct advantage, enabling more reliable
identification of agreement-enabled articles compared to Scopus and Web
of Science, which require reconciliation with proprietary organization
identifiers. Future studies based on the Web of Science will benefit of
the recent integration of ROR-IDs, announced by the end of 2024.

Comparing scholarly databases reveals discrepancies in open access
evidence. Not all publishers deposit Creative Commons (CC) licenses with
Crossref, but comparing it with Unpaywall (integrated into Scopus and
Web of Science) exposes these gaps, as these discovery services also
analyze publishers' websites.

Comparing scholarly databases helps to identify discrepancies in open
access evidence. Crucially, not all publishers deposit CC licences with
Crossref, a limitation that becomes apparent when contrasting Crossref
with Unpaywall's data in Scopus and Web of Science. While Unpaywall can
detect such gaps by parsing publishers' websites for open access
licences, it nevertheless missed a substantial number of CC-licensed
open access articles from Wiley journals indexed in Crossref, likely due
to parsing errors on the publisher's website. This resulted in fewer
open access articles being recorded in Scopus and Web of Science.
Further inconsistencies could be observed between Scopus and Web of
Science, despite both relying on Unpaywall: Web of Science erroneously
labels Elsevier's delayed OA as hybrid, whereas Scopus correctly
categorises it. ONotably, Scopus and Crossref showed greater alignment
than observed in a related comparison between Scopus and OpenAlex
(Alperin et al., 2024). This likely reflects the use of a dedicated list
to identify hybrid open access journals, rather than relying solely on
article-level tags. Given these variations, which may compromise
comparability, research and monitoring exercises should avoid reliance
on a single source. Instead, the selection of open access evidence
should be cross-verified using multiple sources (Huang et al., 2020).
Incorporating expert-curated journal lists can also help prevent
misclassifications based on business models (Visser et al., 2021).

@@

\bibitem[\citeproctext]{ref-Hottenrott_2021}
Hottenrott, H., Rose, M. E., \& Lawson, C. (2021). The rise of multiple
institutional affiliations in academia. \emph{Journal of the Association
for Information Science and Technology}, \emph{72}(8), 1039--1058.
\url{https://doi.org/10.1002/asi.24472}

@@

\bibitem[\citeproctext]{ref-Maddi_2024}
Maddi, A., \& Silva, J. A. T. da. (2024). Beyond authorship: Analyzing
contributions in PLOS ONE and the challenges of appropriate attribution.
\emph{Journal of Data and Information Science}, \emph{9}(3), 88--115.
\url{https://doi.org/10.2478/jdis-2024-0015}
